---
title: "Simulation & Results"
format: html
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(palmerpenguins)
library(epiworldRcalibrate)
library(ggplot2)
library(dplyr)
library(zoo)
library(scales)
library(epiworldR)

```

```{r}
#| include: false
#| message: false
#| warning: false

suppressMessages({
  suppressWarnings({
    source("intro.R")
    source("age_distribution.R")
    source("3_mixing_matrix.R")
    source("4_build_model.R")
  })
})

```

## Real Outbreak Visualization

Lets use LSTM to get the transmission rate. To serve the purpose, I am using roll7, instead of Daily.Cases as the 7 rolling average eliminates any abrupt incresaes in cases as we saw on May 19. Since we are taking 7 day average, I just assigned NULL values to first 7 days. Using calubrate_sir function, population size being 3.3M, recovery rate = 1/7, I predict the transmission_rate.

```{r}
lstm_predictions <- readRDS("lstm_predictions.rds")

cat(
  "===== LSTM CALIBRATION =====\n",
  "Transmission rate (ptran):", lstm_predictions[["ptran"]], "\n",
  "Contact rate (crate):     ", lstm_predictions[["crate"]], "\n",
  "R0:                       ", lstm_predictions[["R0"]], "\n"
)
```

Although we got transmission rate to be 0.017, I just want to check applying various transmission rate to see which one suits best.

We start with single seed - one person infected at start, and below I am trying to find out which transmission rate. Length of period being 61 days, to match actual data and replicating 100 times, just to make it faster. So, initializing transmission rates - c(0.10, 0.15, 0.20, 0.25, 0.30), I built a 100k agent model, 4 incubation days, 1/7 recovery rate and ran it for 100 simulations, for 61 days.

Then, scaled utah cases down to 100k model scale and compared simulated vs scaled real, and picked up the rate which gave lowest RMSE value.

```{r}
# load pre-computed sweep results
sweep_df  <- readRDS("sweep_df.rds")
best_rate <- readRDS("best_rate.rds")

cat("Best transmission rate:", best_rate, "\n")

sweep_df |>
  kableExtra::kable(caption = "Parameter Sweep — Transmission Rate vs Fit") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) |>
  kableExtra::row_spec(
    which.min(sweep_df$RMSE),
    bold = TRUE, background = "#d1e7ff"
  )
```

As shown, LSTM gave much lesser transmission_rate as it was trained on SIR model with homogeneous mixing, starting point being around 30 - 50 cases. Where as, in the above model, I used the contact matrix (age-structured mixing) with single seed. So, I'll go with transmission rate = 0.25.

Now, the model assumes the following points :

1.  I only tested for the above transmission rates, so any if other transmission rate would give a better performance, isn't tested. But since RMSE values decrease upto 0.864, then rose, best transmission rate would be around 0.25 - 0.29
2.  Contact matrix is universal, which is applied to Salt Lake county as well
3.  No vaccination tool added.
4.  No lockdown effects.
5.  Virus effected 1 person only at start, then started to spread.
6.  We did not consider any other viruses like Flu
7.  Person is infected until 7 days. Then recovers.
8.  SEIR model assumes recovered agents are immune thereafter.

```{r}
entities <- lapply(seq_along(dnames), function(i) {
  entity(dnames[i], entity_sizes[i], as_proportion = FALSE)
})

covid_model <- ModelSEIRMixing(
  name              = "COVID-19 Salt Lake County",
  n                 = N_agents,
  prevalence        = 1 / N_agents,
  transmission_rate = best_rate,   # data-driven
  recovery_rate     = 1 / 7,
  incubation_days   = 4,
  contact_matrix    = mm_stochastic
)
```

As shown below, the table displays proportion of age group and their population among 100k agents.

```{r}
# add age group entities
for (e in entities) add_entity(covid_model, e)

print(covid_model)

cat("\n===== MODEL BUILD =====\n")
cat("Total agents:", format(N_agents, big.mark = ","), "\n")
cat("Number of age group entities:", length(entities), "\n")
cat("\nAgent allocation per age group:\n")
print(data.frame(
  age_group    = dnames,
  census_pop   = age_tbl$population,
  fraction     = round(pop_fractions, 4),
  agents       = entity_sizes
))
cat("\nModel summary:\n")
print(covid_model)

```

```{r}

p10 <- data.frame(
  age_group = dnames,
  agents    = entity_sizes
) |>
  mutate(age_group = factor(age_group, levels = dnames)) |>
  ggplot(aes(x = age_group, y = agents)) +
  geom_col(fill = "#2166ac", alpha = 0.85) +
  geom_text(aes(label = scales::comma(agents)),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Agent Allocation by Age Group (Scaled to 100K)",
       x = "Age group", y = "Number of agents") +
  theme_minimal()
print(p10)

```

The model is run for 100 days with 500 simulations.

202,000 rows because, 500 ( replicates ) \* 101 ( 0 to 100 days ) \* 4 ( states - Susceptible, Exposed, Infected, Recovered )

```{r}
saver <- make_saver("total_hist")

run_multiple(
  covid_model,
  ndays    = 100,
  nsims    = 500,
  saver    = saver,
  nthreads = parallel::detectCores(logical = FALSE)
)

sim_results <- run_multiple_get_results(covid_model)
sim_hist    <- sim_results$total_hist

cat(
  "\n===== SIMULATION RUN =====\n",
  "Replicates completed: 500\n",
  "Days simulated: 100\n",
  "Dimensions of total_hist:", paste(dim(sim_hist), collapse = " x "), "\n",
  "Columns:", paste(names(sim_hist), collapse = ", "), "\n",
  "States recorded:", paste(unique(sim_hist$state), collapse = ", "), "\n",
  "\nFirst few rows:\n",
  paste(capture.output(print(head(sim_hist, 12))), collapse = "\n"), "\n"
)

```

The results show increasing cases, starting from 1 case. Its because of the transmission rate. 0.25 is too high. Model chose 0.25 because it minimized RMSE against scaled real data.

R0 = 0.25 \* 7 = 1.75. It means, for a period of 7 days ( until the virus lasts inside a person's body ) he infects 1.75 more people. It is \> 1. Again, depending on the contact matrix, the disease spreads and people get infected.

The real data graph looked like transmission rate was too low, because the cases remained constant each day. I tried using same transmission rate, but since seed = 1, cases remained almost == 1 or 0, each passing day.

One way to stabilize this is to add immunity to people. if 30 - 40% of population were immure to disease, spread wouldn't be so rapid.

```{r}
# ── simulated daily new cases ─────────────────────────────────────────────────
simulated_daily <- sim_hist |>
  dplyr::filter(state == "Recovered") |>
  dplyr::arrange(sim_num, date) |>
  dplyr::group_by(sim_num) |>
  dplyr::mutate(new_cases = counts - lag(counts, default = 0)) |>
  dplyr::ungroup() |>
  dplyr::group_by(date) |>
  dplyr::summarise(
    mean = mean(new_cases),
    lo95 = quantile(new_cases, 0.025),
    hi95 = quantile(new_cases, 0.975),
    .groups = "drop"
  )

# ── identify outbreak vs extinction runs ──────────────────────────────────────
run_outcomes <- sim_hist |>
  dplyr::filter(state == "Recovered", date == max(date)) |>
  dplyr::mutate(
    outcome = ifelse(counts > 1000, "Major outbreak", "Extinction")
  )

cat(
  "===== RUN OUTCOMES =====\n",
  "Extinction runs:    ", sum(run_outcomes$outcome == "Extinction"), "\n",
  "Major outbreak runs:", sum(run_outcomes$outcome == "Major outbreak"), "\n",
  "P(extinction):      ", round(mean(run_outcomes$outcome == "Extinction") * 100, 1), "%\n"
)

# ── use only outbreak runs for comparison ─────────────────────────────────────
outbreak_runs <- run_outcomes |>
  dplyr::filter(outcome == "Major outbreak") |>
  dplyr::pull(sim_num)

simulated_daily_outbreaks <- sim_hist |>
  dplyr::filter(
    state == "Recovered",
    sim_num %in% outbreak_runs
  ) |>
  dplyr::arrange(sim_num, date) |>
  dplyr::group_by(sim_num) |>
  dplyr::mutate(new_cases = counts - lag(counts, default = 0)) |>
  dplyr::ungroup() |>
  dplyr::group_by(date) |>
  dplyr::summarise(
    mean = mean(new_cases),
    lo95 = quantile(new_cases, 0.025),
    hi95 = quantile(new_cases, 0.975),
    .groups = "drop"
  )

# ── scale simulation UP to Utah population ────────────────────────────────────
scale_up <- 3300000 / 100000

simulated_plot <- simulated_daily_outbreaks |>
  mutate(
    mean = mean * scale_up,
    lo95 = lo95 * scale_up,
    hi95 = hi95 * scale_up
  )

# ── real data ─────────────────────────────────────────────────────────────────
real_raw <- covid_data |>
  mutate(date = as.numeric(Date - min(Date))) |>
  select(date, scaled_cases = Daily.Cases)

real_smooth <- covid_data |>
  mutate(date = as.numeric(Date - min(Date))) |>
  select(date, scaled_cases = roll7_filled)

# ── comparison using smoothed real data ───────────────────────────────────────
comparison <- real_smooth |>
  dplyr::inner_join(simulated_plot, by = "date") |>
  dplyr::mutate(
    residual      = scaled_cases - mean,
    abs_error     = abs(residual),
    squared_error = residual ^ 2,
    pct_error     = abs(residual) / (scaled_cases + 1) * 100,
    inside_ci     = scaled_cases >= lo95 & scaled_cases <= hi95
  )

# ── accuracy metrics ──────────────────────────────────────────────────────────
mae         <- mean(comparison$abs_error)
rmse        <- sqrt(mean(comparison$squared_error))
mape        <- mean(comparison$pct_error)
corr        <- cor(comparison$scaled_cases, comparison$mean)
ci_coverage <- mean(comparison$inside_ci) * 100

cat(
  "===== MODEL ACCURACY (outbreak runs only) =====\n",
  sprintf("MAE  (Mean Absolute Error):       %.2f\n", mae),
  sprintf("RMSE (Root Mean Squared Error):   %.2f\n", rmse),
  sprintf("MAPE (Mean Abs Pct Error):        %.1f%%\n", mape),
  sprintf("Correlation (r):                  %.3f\n", corr),
  sprintf("CI Coverage (real inside 95%% CI): %.1f%%\n", ci_coverage)
)

# ── accuracy table ────────────────────────────────────────────────────────────
comparison |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename(
    Day              = date,
    `Real (smoothed)`= scaled_cases,
    `Sim mean`       = mean,
    `Sim 2.5%`       = lo95,
    `Sim 97.5%`      = hi95,
    Residual         = residual,
    `Abs error`      = abs_error,
    `Inside 95% CI`  = inside_ci
  ) |>
  kableExtra::kable(
    caption     = "Actual vs Simulated — Day by Day (Outbreak Runs Only, Days 0–60)",
    format.args = list(big.mark = ",")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed")
  ) |>
  kableExtra::row_spec(
    which(!comparison$inside_ci),
    background = "#fff3cd"
  ) |>
  kableExtra::scroll_box(height = "400px")

# ── plot ──────────────────────────────────────────────────────────────────────
ggplot() +
  geom_ribbon(
    data = simulated_plot,
    aes(x = date, ymin = lo95, ymax = hi95),
    fill = "#d73027", alpha = 0.2
  ) +
  geom_line(
    data = simulated_plot,
    aes(x = date, y = mean, colour = "Simulated"),
    linewidth = 1.1
  ) +
  geom_point(
    data = real_raw,
    aes(x = date, y = scaled_cases, colour = "Real (raw)"),
    size = 1.5, alpha = 0.4
  ) +
  geom_line(
    data = real_smooth,
    aes(x = date, y = scaled_cases, colour = "Real (7-day avg)"),
    linewidth = 1.1, linetype = "dashed"
  ) +
  geom_vline(
    xintercept = 60,
    linetype   = "dotted",
    colour     = "grey50",
    linewidth  = 0.8
  ) +
  annotate(
    "text",
    x      = 62,
    y      = max(real_raw$scaled_cases) * 1.1,
    label  = "Real data ends",
    colour = "grey50",
    size   = 3.5,
    hjust  = 0
  ) +
  scale_x_continuous(
    breaks = seq(0, 100, by = 10),
    limits = c(0, 100)
  ) +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian(ylim = c(0, max(real_raw$scaled_cases) * 1.3)) +  # fix: only one coord_cartesian
  scale_colour_manual(
    values = c(
      "Simulated"        = "#d73027",
      "Real (raw)"       = "#92c5de",
      "Real (7-day avg)" = "#2166ac"
    ),
    name = NULL
  ) +
  labs(
    title    = "Actual vs Simulated Daily Cases",
    subtitle = sprintf(
      "MAE = %.2f | RMSE = %.2f | r = %.3f | CI coverage = %.1f%% | Outbreak runs = %d/500",
      mae, rmse, corr, ci_coverage, length(outbreak_runs)
    ),
    x       = "Day",
    y       = "Daily cases",
    caption = paste(
      "Solid blue = 7-day rolling average | Faint dots = raw daily cases",
      "| Red = simulated mean (outbreak runs only)",
      "| Shaded = 95% CI"
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(colour = "grey40", size = 10)
  )
```

Cause I dont like the graph, lets just try with 40% people with prior immunity and 60% of people vaccinated.

```{r}

prior_immunity <- 0.20   # 40% already immune at start


entities_immune <- lapply(seq_along(dnames), function(i) {
  entity(dnames[i], entity_sizes[i], as_proportion = FALSE)
})

covid_model_immune <- ModelSEIRMixing(
  name              = "COVID-19 SLC + Prior Immunity",
  n                 = N_agents,
  prevalence        = 1 / N_agents,        # single seed
  transmission_rate = best_rate,            # same rate from sweep
  recovery_rate     = 1 / 7,
  incubation_days   = 4,
  contact_matrix    = mm_stochastic
)

for (e in entities_immune) add_entity(covid_model_immune, e)


prior_immunity_tool <- tool(
  name                     = "Prior Immunity",
  susceptibility_reduction = 1.0,   # fully immune
  transmission_reduction   = 0.0,
  recovery_enhancer        = 0.0,
  death_reduction          = 0.0,
  prevalence               = prior_immunity,   # 40% of agents
  as_proportion            = TRUE
)

add_tool(covid_model_immune, prior_immunity_tool)

print(covid_model_immune)


```

```{r}
saver_immune <- make_saver("total_hist")

run_multiple(
  covid_model_immune,
  ndays    = 100,
  nsims    = 500,
  saver    = saver_immune,
  nthreads = parallel::detectCores(logical = FALSE)
)

sim_hist_immune <- run_multiple_get_results(covid_model_immune)$total_hist

# quick check
cat(
  "===== IMMUNITY MODEL CHECK =====\n",
  "Max infected:", max(sim_hist_immune$counts[sim_hist_immune$state == "Infected"]), "\n",
  "Max recovered:", max(sim_hist_immune$counts[sim_hist_immune$state == "Recovered"]), "\n"
)

```

```{r}

get_summary <- function(sh, model_name) {
  sh |>
    dplyr::filter(state == "Recovered") |>
    dplyr::arrange(sim_num, date) |>
    dplyr::group_by(sim_num) |>
    dplyr::mutate(new_cases = counts - lag(counts, default = 0)) |>
    dplyr::ungroup() |>
    dplyr::group_by(date) |>
    dplyr::summarise(
      mean = mean(new_cases),
      lo95 = quantile(new_cases, 0.025),
      hi95 = quantile(new_cases, 0.975),
      .groups = "drop"
    ) |>
    mutate(model = model_name)
}

baseline_summary <- get_summary(sim_hist,        "Baseline (no immunity)")
immune_summary   <- get_summary(sim_hist_immune,  "40% Prior immunity")

# scale up
scale_up <- 3300000 / 100000

baseline_plot <- baseline_summary |>
  mutate(mean = mean * scale_up, lo95 = lo95 * scale_up, hi95 = hi95 * scale_up)

immune_plot <- immune_summary |>
  mutate(mean = mean * scale_up, lo95 = lo95 * scale_up, hi95 = hi95 * scale_up)

all_models <- bind_rows(baseline_plot, immune_plot)


```

```{r}
real_smooth <- covid_data |>
  mutate(date = as.numeric(Date - min(Date))) |>
  select(date, scaled_cases = roll7_filled)

real_raw <- covid_data |>
  mutate(date = as.numeric(Date - min(Date))) |>
  select(date, scaled_cases = Daily.Cases)

ggplot() +
  # baseline CI
  geom_ribbon(
    data = baseline_plot,
    aes(x = date, ymin = lo95, ymax = hi95),
    fill = "#d73027", alpha = 0.15
  ) +
  # immune CI
  geom_ribbon(
    data = immune_plot,
    aes(x = date, ymin = lo95, ymax = hi95),
    fill = "#4393c3", alpha = 0.15
  ) +
  # baseline mean
  geom_line(
    data = baseline_plot,
    aes(x = date, y = mean, colour = "Baseline (no immunity)"),
    linewidth = 1.1
  ) +
  # immune mean
  geom_line(
    data = immune_plot,
    aes(x = date, y = mean, colour = "40% Prior immunity"),
    linewidth = 1.1
  ) +
  # real raw dots
  geom_point(
    data = real_raw,
    aes(x = date, y = scaled_cases, colour = "Real (raw)"),
    size = 1.5, alpha = 0.4
  ) +
  # real smooth line
  geom_line(
    data = real_smooth,
    aes(x = date, y = scaled_cases, colour = "Real (7-day avg)"),
    linewidth = 1.1, linetype = "dashed"
  ) +
  geom_vline(
    xintercept = 60,
    linetype   = "dotted",
    colour     = "grey50",
    linewidth  = 0.8
  ) +
  annotate(
    "text",
    x = 62, y = max(real_raw$scaled_cases) * 1.1,
    label = "Real data ends",
    colour = "grey50", size = 3.5, hjust = 0
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10), limits = c(0, 100)) +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian(ylim = c(0, max(real_raw$scaled_cases) * 1.5)) +
  scale_colour_manual(
    values = c(
      "Baseline (no immunity)" = "#d73027",
      "40% Prior immunity"     = "#228B22",
      "Real (raw)"             = "#92c5de",
      "Real (7-day avg)"       = "#2166ac"
    ),
    name = NULL
  ) +
  labs(
    title    = "Baseline vs Prior Immunity Model — Daily Cases",
    subtitle = "Salt Lake County | 100K agents | 500 replicates | Single seed",
    x        = "Day",
    y        = "Daily cases",
    caption  = paste(
      "Red = baseline (0% immunity) | Blue = 40% prior immunity",
      "| Dashed = real 7-day avg | Shaded = 95% CI",
      "| Simulation scaled up ×33 to Utah population"
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(colour = "grey40", size = 10)
  )

```

```{r}
get_accuracy <- function(sim_plot, model_name) {
  comp <- real_smooth |>
    dplyr::inner_join(sim_plot, by = "date") |>
    dplyr::mutate(
      abs_error     = abs(scaled_cases - mean),
      squared_error = (scaled_cases - mean)^2,
      inside_ci     = scaled_cases >= lo95 & scaled_cases <= hi95
    )
  
  data.frame(
    Model       = model_name,
    MAE         = round(mean(comp$abs_error), 2),
    RMSE        = round(sqrt(mean(comp$squared_error)), 2),
    Correlation = round(cor(comp$scaled_cases, comp$mean), 3),
    CI_Coverage = round(mean(comp$inside_ci) * 100, 1)
  )
}

accuracy_comparison <- rbind(
  get_accuracy(baseline_plot, "Baseline (no immunity)"),
  get_accuracy(immune_plot,   "40% Prior immunity")
)

accuracy_comparison |>
  kableExtra::kable(caption = "Model Comparison — Baseline vs Prior Immunity") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) |>
  kableExtra::row_spec(
    which.min(accuracy_comparison$RMSE),
    bold = TRUE, background = "#d1e7ff"
  ) |>
  kableExtra::row_spec(0, bold = TRUE,
                       background = "#343a40", color = "white")
```
